Estimating duration in the presence of misclassification

### 2024-07-29 ATACCC.tex
Rob second p75 comment
Rob typo comments
Christl acronym comment

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient implementation of the NUTS algorithm (see \cref{E-MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model}):
\begin{align}
v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $t$, is continuous and defined for individual $i$ such that $i$s first test is at $t = 0$.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert t - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(t-\tau_{\max,i})$ is small and hence $a\exp(b(t-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(t-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for representative viral load trajectories).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.

@@

Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$.

@@

The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.

The two models give very similar estimates for the population-level parameters (see \cref{ATACCC:fig:compare-hakki-modified}).
These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
As the estimates are similar and the modified model converges better, I will use only the results from the modified model.
\begin{figure}
  \centering \includegraphics{ATACCC/compare_hakki_modified}
  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$, and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}). Mean only shown for the unvaccinated group, the primary group of interest. The groups share the covariance matrix. These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}), have very similar estimates between the models. \label{ATACCC:fig:compare-hakki-modified}}
\end{figure}

>>>>

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient implementation of the NUTS (the No U-Turn Sampler) algorithm (see \cref{E-MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model}):
\begin{align}
v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{\max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $t$, is continuous and defined for individual $i$ such that $i$s first test is at $t = 0$.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert t - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(t-\tau_{\max,i})$ is small and hence $a\exp(b(t-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(t-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for representative viral load trajectories).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.

@@

Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
One sample is produced for each iteration of the MCMC (Markov Chain Monte Carlo) output used to estimate $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$.

@@

The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS (Effective Sample Size) < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.

The two models give very similar estimates for the population-level parameters (see \cref{ATACCC:fig:compare-hakki-modified}).
These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
As the estimates are similar and the modified model converges better, I will use only the results from the modified model.
In this study, $\vlod = 0$.
\begin{figure}
  \centering \includegraphics{ATACCC/compare_hakki_modified}
  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$, and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}). Mean only shown for the unvaccinated group, the primary group of interest. The groups share the covariance matrix. These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}), have very similar estimates between the models. \label{ATACCC:fig:compare-hakki-modified}}
\end{figure}

### 2024-07-29 biology-data.tex
Christl acronym comment

SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
Coronaviruses are RNA (Ribonucleic acid) viruses, which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.

@@

\begin{figure}
  \centering \includegraphics[width=\textwidth]{biology-data/CIS-positivity}
  \caption[CIS positivity]{%
    CIS positivity rate in the period considered in this thesis (Sep 2020 until Jan 2021).
    Dots show the unadjusted positivity rate (days with fewer than 100 tests are excluded for readability).
    Ribbon shows modelled 95\% CrI for the positivity, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
  }
  \label{biology-data:fig:CIS-positivity}
\end{figure}

@@

For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
The SRS is a TRE (Trusted Research Environment) that allows researchers to access sensitive data in a secure environment.
The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.

>>>>

SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV (Severe Acute Respiratory Syndrome--related CoronaVirus) and MERS-CoV (Middle East Respiratory Syndrome--related CoronaVirus).
Coronaviruses are RNA (Ribonucleic acid) viruses, which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.

@@

\begin{figure}
  \centering \includegraphics[width=\textwidth]{biology-data/CIS-positivity}
  \caption[CIS positivity]{%
    CIS positivity rate in the period considered in this thesis (Sep 2020 until Jan 2021).
    Dots show the unadjusted positivity rate (days with fewer than 100 tests are excluded for readability).
    Ribbon shows modelled 95\% CrI (Credible Interval) for the positivity, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
  }
  \label{biology-data:fig:CIS-positivity}
\end{figure}

@@

For privacy reasons, the CIS data are stored in the ONS's SRS~\autocite{onsSRS}.
The SRS is a TRE (Trusted Research Environment) that allows researchers to access sensitive data in a secure environment.
The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.

### 2024-07-29 incidence-prevalence.tex
Christl acronym comment

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

@@

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.

@@

Reproduction numbers in models with structured populations are more complex than the homogeneous case.
The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{a'a}$. 
However, how to average these contributions to calculate the basic reproduction number is not obvious.

>>>>

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cdf (cumulative density function), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

@@

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (Who Acquires Infection From Whom) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.

@@

Reproduction numbers in models with structured populations are more complex than the homogeneous case.
The WAIFW matrix allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{a'a}$. 
However, how to average these contributions to calculate the basic reproduction number is not obvious.

### 2024-07-29 introduction.tex
Christl acronym comment

Infection with the SARS-CoV-2 virus is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
Beyond the lives lost, there were massive health, economic, and societal costs.

@@

Data on the severe consequences of an infection are typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

@@

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf of a distribution, or a transformation of it.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

@@

Many longitudinal studies were conducted throughout the pandemic.
They aimed to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus.
These include the CIS, previously mentioned.
% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.

Another study, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
ATACCC recruited individuals from contacts of known infections reported to NHS Test and Trace, and collected data on them for 20 days.
This included daily RT-PCR tests, which allowed quantification of their viral load (the amount of virus in an individual), a biomarker.
The density of the sampling allows for modelling the trajectory of viral load to estimate the duration of detectability (how long an individual is RT-PCR positive for).
However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.

>>>>

Infection with the SARS-CoV-2 (Severe Acute Respiratory Syndrome CoronaVirus 2) virus is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
Beyond the lives lost, there were massive health, economic, and societal costs.

@@

Data on the severe consequences of an infection are typically less affected by biases.
An extreme example is the development of AIDS (Acquired ImmunoDeficiency Syndrome) following infection by HIV (Human Immunodeficiency Virus).
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

@@

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf (probability density function) of a distribution, or a transformation of it.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

@@

Many longitudinal studies were conducted throughout the pandemic.
They aimed to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV-2 virus.
These include the CIS, previously mentioned.
% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.

Another study, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
ATACCC recruited individuals from contacts of known infections reported to NHS (the National Health Service) Test and Trace, and collected data on them for 20 days.
This included daily RT-PCR tests, which allowed quantification of their viral load (the amount of virus in an individual), a biomarker.
The density of the sampling allows for modelling the trajectory of viral load to estimate the duration of detectability (how long an individual is RT-PCR positive for).
However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.
